---
title: "Complete Margot Workflow: From Data to Interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Margot Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates the complete margot workflow for causal inference, based on the teaching materials from PSYCH 434. The workflow consists of three main stages:

1. **Data Preparation**: Converting longitudinal data to wide format
2. **Causal Inference**: Estimating treatment effects using causal forests
3. **Interpretation**: Visualising and understanding results

## Installation

The margot package has a modular design. Install only what you need:

```{r}
# core package (data manipulation and basic functions)
install.packages("margot")

# for causal inference models (optional)
install.packages(c("grf", "lmtp", "policytree", "maq"))

# for visualisation (optional)
install.packages(c("ggplot2", "patchwork", "ggokabeito"))

# for reporting tables (optional)
install.packages(c("gt", "gtsummary", "flextable"))

# or install everything at once (future)
# install.packages("margot.models")  # all estimation packages
# install.packages("margot.viz")     # all visualisation packages
# install.packages("margot.report")  # all reporting packages
```

## Stage 1: Data Preparation

### Loading and Exploring Data

```{r}
library(margot)
library(dplyr)

# the package includes example data
data(df_nz)

# define variable groups
baseline_vars <- c(
  "male", "age", "eth_cat", "partner", "agreeableness",
  "conscientiousness", "extraversion", "honesty_humility", 
  "openness", "neuroticism", "sample_weights"
)

exposure_var <- "forgiveness"

outcome_vars <- c(
  "alcohol_frequency", "alcohol_intensity", 
  "hours_exercise", "hours_work", "ego_rubin"
)
```

### Converting to Wide Format

The `margot_wide_machine()` function handles the complete data preparation pipeline:

```{r}
# prepare data for causal inference
wide_data <- margot_wide_machine(
  data = df_nz,
  baseline_vars = baseline_vars,
  exposure_var = exposure_var,
  outcome_vars = outcome_vars
)

# check the structure
str(wide_data)
```

## Stage 2: Causal Inference

### Estimating Treatment Effects

```{r}
# run causal forest (requires grf package)
results <- margot_causal_forest(
  data = wide_data,
  exposure = exposure_var,
  outcomes = outcome_vars,
  baseline_vars = baseline_vars,
  weights = "sample_weights"
)
```

### Screening for Heterogeneity

```{r}
# identify which outcomes show treatment effect heterogeneity
heterogeneity_results <- margot_rate(results)
```

### Policy Learning

```{r}
# learn optimal treatment policies
policy_results <- margot_policy(
  results,
  outcomes = outcome_vars,
  baseline_vars = baseline_vars
)
```

## Stage 3: Interpretation and Visualisation

### Visualising Treatment Effects

```{r}
# plot average treatment effects (requires ggplot2)
margot_plot(
  results,
  type = "effects",
  title = "Average Treatment Effects of Forgiveness"
)

# create table output (requires gt)
margot_plot(
  results,
  type = "table",
  format = "publication"
)
```

### Understanding Heterogeneity

```{r}
# visualise policy trees
margot_plot_policy_tree(
  policy_results,
  outcome = "ego_rubin"
)

# plot qini curves
margot_plot_qini(
  policy_results,
  outcome = "ego_rubin"
)
```

## Working with Missing Packages

If you haven't installed optional packages, margot provides helpful error messages:

```{r, eval = TRUE, error = TRUE}
# example: trying to use causal forest without grf installed
# margot_causal_forest(wide_data)
# Error: Package 'grf' is required for margot_causal_forest() (causal forest estimation).
# Install it with: install.packages('grf')
# For all estimation packages: install.packages('margot.models')
```

## Complete Example Script

Here's a minimal complete workflow:

```{r}
# load packages
library(margot)
library(dplyr)

# prepare data
data(df_nz)

# define variables
baseline_vars <- c("male", "age", "partner")
exposure_var <- "forgiveness"
outcome_vars <- c("ego_rubin")

# run complete pipeline
wide_data <- margot_wide_machine(
  df_nz, baseline_vars, exposure_var, outcome_vars
)

# estimate effects (requires grf)
if (requireNamespace("grf", quietly = TRUE)) {
  results <- margot_causal_forest(
    wide_data, exposure_var, outcome_vars, baseline_vars
  )
  
  # visualise (requires ggplot2)
  if (requireNamespace("ggplot2", quietly = TRUE)) {
    margot_plot(results)
  }
}
```

## Further Resources

- Full course materials: https://go-bayes.github.io/psych-434-2025/
- Package documentation: https://go-bayes.github.io/margot/
- GitHub repository: https://github.com/go-bayes/margot