---
title: "Average-effect reporting and outcome scales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Average-effect reporting and outcome scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(margot)
```

`margot_plot_ate()`, `margot_table_ate()` and `margot_interpret_ate()` produce a figure, a numerical table and prose from the same reporting calculation. `margot_plot()` remains available and returns all three as `plot`, `transformed_table` and `interpretation`. The example below uses invented estimates and sensitivity quantities; it fits no model and uses no participant data.

```{r}
estimates <- data.frame(
  outcome = c("wellbeing_z", "alcohol_z"),
  ATE = c(0.15, -0.10),
  E_Value = c(1.8, 1.6),
  E_Val_bound = c(1.3, 1.2)
)
estimates[["2.5 %"]] <- c(0.04, -0.25)
estimates[["97.5 %"]] <- c(0.28, -0.02)
scales <- data.frame(
  outcome = c("wellbeing_z", "alcohol_z"),
  transformation = "identity",
  center = c(4, 3), scale = c(1.2, 2), orientation = 1,
  unit = c("points", "drinks"), unit_multiplier = 1
)
margot_plot_ate(estimates, scale_info = scales, e_val_bound_threshold = 1)
margot_table_ate(estimates, scale_info = scales, e_val_bound_threshold = 1)
cat(margot_interpret_ate(estimates, scale_info = scales, e_val_bound_threshold = 1))
```

The graph retains the supplied model-scale estimates and intervals. The table retains those numbers and adds `reported_estimate`, `reported_lower`, `reported_upper`, `reporting_quantity` and `reporting_unit`. Its `confidence_level` column preserves each interval's supplied coverage. Prose reports the supported conversion alongside the model-scale estimate. The E-value threshold controls which estimates enter the prose; it does not establish identification or remove the need to report other planned outcomes. E-values remain tied to their supplied scientific scale and calculation, not to a change of display units.

## Record preparation constants

Let `Y` denote the measured outcome and `g` its declared transformation. Metadata describes the model outcome as `orientation * (g(Y) - center) / scale`. Each row of `scale_info` matches an original outcome key before display labels change. Store the constants from preparation rather than recomputing them from a reporting subset.

| Field | Meaning | Default |
|---|---|---|
| `outcome` | Original outcome key in the estimate table | Required |
| `transformation` | `identity`, `log`, or `log1p` | `identity` |
| `center` | Centre subtracted during preparation | 0 |
| `scale` | Positive divisor used during preparation | 1 |
| `orientation` | Outcome orientation, 1 or -1 | 1 |
| `unit` | Unit label for the reported affine difference | Empty |
| `unit_multiplier` | Positive affine unit conversion, such as 60 for hours to minutes | 1 |

For an affine mean difference, the centre cancels. The estimate and both supplied confidence limits are multiplied by `orientation * scale * unit_multiplier`; reversed limits are reordered. This preserves asymmetric intervals and numerical precision. Display functions may round printed numbers, but the numerical table retains their precision. A ratio of centred outcomes cannot generally be converted to an original-scale ratio from that contrast alone; unsupported ratio conversions fail explicitly.

## The applied NZAVS convention uses unlogged values before standardisation

For the New Zealand Attitudes and Values Study (NZAVS), the current applied preparation convention keeps continuous baseline and lagged adjustment variables on their unlogged scientific scales before z-standardisation. Exposure columns used to define an intervention retain the registered exposure scale. Terminal outcomes also remain unlogged: the exercise outcome is `hours_exercise`, measured in hours per week. Scoring and orientation remain explicit measurement decisions; omitting a logarithm does not remove those decisions.

For each terminal outcome, preparation records the unweighted mean and sample standard deviation among that outcome's primary estimation rows. It applies those constants once and retains them across primary fits, sensitivity analyses, groups, policy folds, and reports. This is the same standardisation convention used by the developing GRF workflow. The reference rows and their missing-data rule belong in the preparation record. An unweighted standardisation reference does not replace the analysis weights used to estimate a target-population contrast. Baseline and lagged adjustment variables retain their own preparation constants, separately from terminal-outcome constants.

The following example uses three invented primary-row exercise values and an invented estimated contrast, interval, and E-values. It fits no model. The saved mean is 2 hours per week, and the sample standard deviation is 2 hours per week.

```{r exercise-affine}
primary_hours_exercise <- c(0, 2, 4)
exercise_center <- mean(primary_hours_exercise)
exercise_scale <- sd(primary_hours_exercise)
exercise_z <- (primary_hours_exercise - exercise_center) / exercise_scale

exercise_estimate <- data.frame(
  outcome = "hours_exercise_z",
  ATE = 0.25,
  `2.5 %` = 0.10,
  `97.5 %` = 0.40,
  confidence_level = 0.95,
  E_Value = 1.8,
  E_Val_bound = 1.4,
  check.names = FALSE
)
exercise_scale_info <- data.frame(
  outcome = "hours_exercise_z",
  transformation = "identity",
  center = exercise_center,
  scale = exercise_scale,
  orientation = 1,
  unit = "hours per week",
  unit_multiplier = 1
)
exercise_table <- margot_table_ate(
  exercise_estimate,
  scale_info = exercise_scale_info,
  e_val_bound_threshold = 1
)
exercise_table[, c("ATE", "reported_estimate", "reported_lower", "reported_upper")]
cat(margot_interpret_ate(
  exercise_estimate,
  scale_info = exercise_scale_info,
  label_mapping = list(hours_exercise_z = "Weekly exercise"),
  e_val_bound_threshold = 1
))
```

The contrast of 0.25 standard deviations corresponds to 0.5 hours per week, with the supplied 95% interval transformed to 0.2–0.8 hours per week. The mean cancels from the difference. These are two scales for the same affine mean contrast; the original-unit companion is not a new fit.

## Logarithms remain available for other outcome definitions

The applied NZAVS convention above does not remove Margot's general-purpose logarithmic reporting. Historical analyses retain their specifications, and another scientific question may explicitly define a logged outcome. The reported quantity must match that definition.

Exponentiating a difference of mean log outcomes gives a ratio of geometric means. With `log1p`, it gives a ratio of geometric means of `Y + 1`. Neither quantity is generally an arithmetic mean difference or an arithmetic mean ratio. A transformed contrast cannot recover a missing arithmetic-mean causal estimand merely by substituting a reference mean.

```{r}
control <- c(0, 3)
policy <- c(1, 7)
log_difference <- mean(log1p(policy)) - mean(log1p(control))
c(shifted_geometric_mean_ratio = exp(log_difference),
  arithmetic_mean_difference = mean(policy) - mean(control))
```

The shifted geometric ratio is 2, whereas the arithmetic mean difference is 2.5. These are different scientific quantities. Nonlinear reported values therefore receive explicit geometric-ratio labels, and legacy arithmetic `_original` columns remain missing. Do not pass the displayed geometric ratio to a risk-ratio E-value formula.

## Existing calls

Calls to `margot_plot()` retain the combined list, argument precedence and saving behaviour. The three focused interfaces select their respective component; `save_output = TRUE` still saves the complete list. Existing `original_df` calls infer transformations from outcome suffixes and matching unstandardised source columns, with a warning that scale constants may be recomputed. Legacy log names mean `log1p`; explicit `scale_info` distinguishes ordinary logarithms. Missing or ambiguous source columns require explicit metadata. No reporting path should substitute an invented population mean.

When regenerating an existing study report, retain its estimation-version record and separately record the reporting version and corrected quantities. Reporting corrections do not imply that an earlier study was estimated with a later package release.
