---
title: "Average-effect reporting and outcome scales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Average-effect reporting and outcome scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(margot)
```

`margot_plot_ate()`, `margot_table_ate()` and `margot_interpret_ate()` produce a figure, a numerical table and prose from the same reporting calculation. `margot_plot()` remains available and returns all three as `plot`, `transformed_table` and `interpretation`. The example below uses invented estimates and sensitivity quantities; it fits no model and uses no participant data.

```{r}
estimates <- data.frame(
  outcome = c("wellbeing_z", "alcohol_z"),
  ATE = c(0.15, -0.10),
  E_Value = c(1.8, 1.6),
  E_Val_bound = c(1.3, 1.2)
)
estimates[["2.5 %"]] <- c(0.04, -0.25)
estimates[["97.5 %"]] <- c(0.28, -0.02)
scales <- data.frame(
  outcome = c("wellbeing_z", "alcohol_z"),
  transformation = "identity",
  center = c(4, 3), scale = c(1.2, 2), orientation = 1,
  unit = c("points", "drinks"), unit_multiplier = 1
)
margot_plot_ate(estimates, scale_info = scales, e_val_bound_threshold = 1)
margot_table_ate(estimates, scale_info = scales, e_val_bound_threshold = 1)
cat(margot_interpret_ate(estimates, scale_info = scales, e_val_bound_threshold = 1))
```

The graph retains the supplied model-scale estimates and intervals. The table retains those numbers and adds `reported_estimate`, `reported_lower`, `reported_upper`, `reporting_quantity` and `reporting_unit`. Its `confidence_level` column preserves each interval's supplied coverage. Prose reports the supported conversion alongside the model-scale estimate. The E-value threshold controls which estimates enter the prose; it does not establish identification or remove the need to report other planned outcomes. E-values remain tied to their supplied scientific scale and calculation, not to a change of display units.

## Record preparation constants

Let `Y` denote the measured outcome and `g` its declared transformation. Metadata describes the model outcome as `orientation * (g(Y) - center) / scale`. Each row of `scale_info` matches an original outcome key before display labels change. Store the constants from preparation rather than recomputing them from a reporting subset.

| Field | Meaning | Default |
|---|---|---|
| `outcome` | Original outcome key in the estimate table | Required |
| `transformation` | `identity`, `log`, or `log1p` | `identity` |
| `center` | Centre subtracted during preparation | 0 |
| `scale` | Positive divisor used during preparation | 1 |
| `orientation` | Outcome orientation, 1 or -1 | 1 |
| `unit` | Unit label for the reported affine difference | Empty |
| `unit_multiplier` | Positive affine unit conversion, such as 60 for hours to minutes | 1 |

For an affine mean difference, the centre cancels. The estimate and both supplied confidence limits are multiplied by `orientation * scale * unit_multiplier`; reversed limits are reordered. This preserves asymmetric intervals and numerical precision. Display functions may round printed numbers, but the numerical table retains their precision. A ratio of centred outcomes cannot generally be converted to an original-scale ratio from that contrast alone; unsupported ratio conversions fail explicitly.

## Logarithms change the reported quantity

Exponentiating a difference of mean log outcomes gives a ratio of geometric means. With `log1p`, it gives a ratio of geometric means of `Y + 1`. Neither quantity is generally an arithmetic mean difference or an arithmetic mean ratio. A transformed contrast cannot recover a missing arithmetic-mean estimand merely by substituting a reference mean.

```{r}
control <- c(0, 3)
policy <- c(1, 7)
log_difference <- mean(log1p(policy)) - mean(log1p(control))
c(shifted_geometric_mean_ratio = exp(log_difference),
  arithmetic_mean_difference = mean(policy) - mean(control))
```

The shifted geometric ratio is 2, whereas the arithmetic mean difference is 2.5. These are different scientific quantities. Nonlinear reported values therefore receive explicit geometric-ratio labels, and legacy arithmetic `_original` columns remain missing. Do not pass the displayed geometric ratio to a risk-ratio E-value formula.

## Existing calls

Calls to `margot_plot()` retain the combined list, argument precedence and saving behaviour. The three focused interfaces select their respective component; `save_output = TRUE` still saves the complete list. Existing `original_df` calls infer transformations from outcome suffixes and matching unstandardised source columns, with a warning that scale constants may be recomputed. Legacy log names mean `log1p`; explicit `scale_info` distinguishes ordinary logarithms. Missing or ambiguous source columns require explicit metadata. No reporting path should substitute an invented population mean.

When regenerating an existing study report, retain its estimation-version record and separately record the reporting version and corrected quantities. Reporting corrections do not imply that an earlier study was estimated with a later package release.
