---
title: "Migrating off `qs`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating off `qs`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## What happened to `qs`

Earlier versions of margot used the `qs` package to write fast, compressed
serialisations of fitted models, plot lists, lmtp results, and other
analysis artefacts. `margot` users may have projects that accumulate archives of `.qs` files.

`qs` was archived from CRAN in 2025 because it depends on R-internal C
symbols (`LEVELS`, `OBJECT`, `ENCLOS`, `Rf_allocSExp`, …) that **R 4.6
removed**. As a result `qs` cannot be built or installed on R 4.6+ — even
from GitHub source.

This is a pain, and a lesson.

margot 1.0.320 deprecates the package's qs-named helpers for new storage.
New projects should use standard RDS save formats wherever possible. You can do this in margot using `here_save()` and `here_read()`. The parquet-backed arrow helpers can be used for large rectangular data (such as dataframes): `here_save_arrow()` and `here_read_arrow()`.  It is unlikely that parquet will change in ways that render old files invalid, however, the loss of `qs` is a reminder to favour robustness where failure is costly.  Direct local
conversion remains available when an older R environment can load `qs`, and
the Docker migration bridge remains available when local `qs` is not possible. So there are ways to get your stored data back.

## What changed in margot

- `here_save_qs(obj, name, dir_path)` and `here_read_qs(name, dir_path)`
  are deprecated. They still handle existing `.qs2` files, but new work
  should use `here_save()` and `here_read()`.
- `here_read_qs()` still reads legacy `.qs` files when optional `qs` is
  installed, so old archives can be opened and migrated without Docker in
  compatible R environments.
- `here_save_arrow(obj, name)` now round-trips arbitrary R objects — not
  just data frames. Lists, fitted models, ggplots, and result objects
  are wrapped in a single-row "margot envelope" parquet.
- `qs` is listed in `Suggests`, not `Imports`. Recall, fresh installs of margot on R 4.6+ *cannot* install it, but older R environments that already
  have `qs` can still use the direct converter.

Pipelines that already have `.qs2` files can still read them through the
deprecated helpers while migrating to arrow. Pipelines with old `.qs` files
need a one-off conversion first.

## Decision tree

```
You have legacy .qs files you need to read.
│
├── Do you have an R 4.5 environment available  ───  yes  ──>  margot_convert_qs_dir(dir)
│   with `qs` already installed?
│
└── Are you on R 4.6+ with no qs?  ───────────────  yes  ──>  margot_convert_qs_dir_docker(dir)
                                                              (needs Docker)
```

Use the direct converter when you can. Use Docker only when local `qs` is not
available.

## Path A: direct local conversion (older machines that still have `qs`)

If you have an older R environment where `qs` already works, use margot's
direct converter there:

```{r}
margot::margot_convert_qs_dir(
  "~/path/to/your/cache/dir"
)
```

This recursively finds `.qs` files, writes `.qs2` siblings with `qs2`, and
read-verifies each converted file before reporting success. Originals are kept
by default.

Once you have spot-checked the converted files, remove the originals with:

```{r}
margot::margot_convert_qs_dir(
  "~/path/to/your/cache/dir",
  delete_qs = TRUE
)
```

The converter verifies the `.qs2` round trip before deleting a source `.qs`
file.

## Path B: conversion via Docker (R 4.6+ users)

If you are on R 4.6+ and `qs` will not install, run the conversion inside
a Docker container. margot does the heavy lifting; you only need a
working Docker engine on your machine.

### Prerequisites at a glance

| | What you need | How to check |
|---|---|---|
| 1 | A working Docker engine | `docker info` shows a "Server" section |
| 2 | `margot` >= 1.0.320 in your R session | `packageVersion("margot") >= "1.0.320"` |
| 3 | `qs2` in your host R session | `requireNamespace("qs2")` |
| 4 | ~2 GB free disk (image + libs + converted files) | – |

`qs2` is a hard `Imports` of margot, so a fresh `install.packages("margot")`
or a `pak::pak("go-bayes/margot")` will pull it automatically. If you
upgraded margot in place from a much older version, run
`install.packages("qs2")` once.

You do **not** need `qs`, R 4.5, Rcpp, or any compiler installed on your
host. Everything that needs `qs` runs inside the container.

### One-time Docker setup

#### macOS

The lightest option is **Colima** (a small Linux VM that the standard
Docker CLI talks to). It is free and CLI-only:

```bash
# install Homebrew first if you do not have it: https://brew.sh
brew install docker colima
colima start --memory 6 --cpu 4
```

Verify Docker is reachable before continuing:

```bash
docker info | head -10
```

You should see lines such as `Server Version: 29.x.x` and
`Operating System: Ubuntu 24.04`. If the output ends with
"failed to connect to the docker API", run `colima start` again.

**Docker Desktop** also works if you prefer a GUI app. Download from
<https://www.docker.com/products/docker-desktop>, install, launch the
app, and wait for the whale icon in the menu bar to settle. Note that
Docker Desktop requires a paid subscription for organisations above
certain size thresholds; Colima has no such restriction.

#### Linux

Most distributions package Docker directly. On Debian/Ubuntu:

```bash
sudo apt-get update
sudo apt-get install docker.io
sudo systemctl start docker
sudo usermod -aG docker $USER   # log out and back in for this to apply
```

On Fedora/RHEL: `sudo dnf install docker`.
Then verify with `docker info`.

#### Windows

Install **Docker Desktop for Windows** (with the WSL 2 backend
recommended): <https://www.docker.com/products/docker-desktop>.
After installation, restart and open Docker Desktop once so the
engine starts. Then verify with `docker info` from a terminal.

### Run the migration

From your normal R 4.6 session:

```{r}
margot::margot_convert_qs_dir_docker(
  "~/path/to/your/cache/dir"
)
```

The first run pulls the `rocker/r-ver:4.5` image (~600 MB) and compiles
`qs` 0.27.3 from a pinned **2024-12-01 Posit Package Manager snapshot**
so the build is reproducible across machines and dates. Subsequent runs
reuse the cached image and library and finish quickly.

Per-file output looks like:

```
ℹ starting docker container (rocker/r-ver:4.5); first run pulls the image and compiles qs.
ℹ mounted: /Users/you/path/to/cache -> /data
found 12 .qs file(s) under /data
ok     project-a/models_binary.qs
ok     project-a/policy_workflow.qs
skip   project-b/old.qs (.qs2 exists)
…
converted: 11, skipped: 1, failed: 0
✔ conversion finished. spot-check a .qs2 file before re-running with delete_qs = TRUE.
```

### Spot-check, then delete

After the first run, read one of the new `.qs2` files in your normal
session to confirm the migration:

```{r}
obj <- qs2::qs_read("~/path/to/your/cache/dir/something.qs2")
str(obj, max.level = 1)
```

Once you're satisfied, run again with `delete_qs = TRUE` to remove the
originals:

```{r}
margot::margot_convert_qs_dir_docker(
  "~/path/to/your/cache/dir",
  delete_qs = TRUE
)
```

The container performs an `identical()` read-verify *before* any
deletion, so the safety net stays in place even with `delete_qs = TRUE`. Still, make certain you want to delete the old qs files before setting this parameter to TRUE.

## Caveats

- **Bind-mount scope.** Colima and Docker Desktop on macOS share `~/`
  into the VM by default. If your archive lives outside `~/` (e.g. on
  an external volume), either move the work into `~/` first or extend
  the Colima mount with `colima start --mount /Volumes/External:w`.
- **Memory.** Default Colima allocates 2 GiB of RAM, which can OOM on
  very large `qs` files (the deserialised object lives in memory).
  `colima start --memory 6 --cpu 4` is a sensible baseline; users with
  multi-gigabyte caches may want more.
- **Rocker image size.** The first pull is ~600 MB. Once cached locally
  it's reused for every subsequent migration run on that machine.

## After migration

Move day-to-day object storage to the arrow helpers:

```{r}
margot::here_save(my_models, "models")
my_models <- margot::here_read("models")
```

LMTP crash-recovery checkpoints are separate. Keep using
`margot_lmtp(save_output = TRUE)` for long LMTP runs; those per-model
checkpoints are written as `.rds` files so interrupted work can be restored
without `qs`, `qs2`, Arrow, or Docker.

## Reporting issues

If `margot_convert_qs_dir_docker()` fails on a specific archive, please
open an issue at <https://github.com/go-bayes/margot/issues> with:

- the per-file `FAIL` line(s) printed by the function,
- the host R version (`R.version.string`),
- the Docker engine in use (`docker info | head -20`),
- the size of the failing `.qs` file (`file.info(path)$size`).

The most common failure mode is an out-of-memory kill inside the
container; the fix is usually `colima stop && colima start --memory 8`
(or higher), then re-run.
