---
title: "Standard GRF and policy-tree workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard GRF and policy-tree workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette describes the standard `margot` workflow for generalised random
forest (GRF) analyses with policy-tree reporting. The workflow is designed for
outcome-wide studies: the same exposure, covariates, and analysis population are
used to estimate effects for multiple outcomes.

The workflow separates four tasks:

1. Estimate average treatment effects (ATEs) for each outcome.
2. Diagnose whether the forest predictions show calibrated heterogeneity.
3. Evaluate policy-tree learning with held-out folds.
4. Summarise cross-outcome recurrence descriptively.

The ATE task estimates the primary causal estimand. The policy-tree task asks
whether a shallow rule can summarise useful variation in the forest's doubly
robust action scores. Because a policy tree is an optimiser, policy-tree
learning should be evaluated on held-out observations using cross-validation.

The standard workflow uses the package defaults unless a registration states an
override. For `margot_policy_tree_cv()`, the defaults compare depths one and two
with five held-out folds repeated 20 times, select depth two only when it clears
the package parsimony rule, and prefer `fastpolicytree` when it is available.

## Reporting convention

Policy trees optimise action-score rewards. The reporting layer therefore
separates the action selected by the tree from the signed score contrast shown
in each leaf.

Let $\Gamma_{ja}$ denote the action score for observation $j$ under action $a$,
let $w_j$ denote the evaluation weight, let $L$ denote a policy-tree leaf, and
let $E_L$ denote the evaluation observations routed to that leaf. For binary
actions $C$ and $T$, `margot` reports the signed evaluation-sample
treatment-control contrast

$$
\Delta_L =
\frac{\sum_{j \in E_L} w_j\{\Gamma_{jT} - \Gamma_{jC}\}}
     {\sum_{j \in E_L} w_j}.
$$

Positive values of $\Delta_L$ favour treatment, negative values favour control,
and zero values indicate no directional contrast in the action scores. The
selected action is reported separately. For a tree fitted on training
observations $S_L$, the stored action is learned as

$$
\pi(L) = \arg\max_{a \in \{C,T\}}
\frac{\sum_{j \in S_L} w_j \Gamma_{ja}}{\sum_{j \in S_L} w_j}.
$$

If action-score means tie within numerical tolerance, the leaf has no
directional contrast. The reported selected action follows the action stored by
the fitted tree or a prespecified tie rule; it should not be read as a
substantive preference when $\Delta_L = 0$. In held-out CV summaries, the
stored action is learned on training folds and $\Delta_L$ is computed on
held-out observations; the held-out table does not reselect actions from
held-out means.

This distinction affects interpretation when every leaf selects the same action.
Such a tree may still split observations because score-contrast magnitude varies
across the covariate space, but it should not be described as a selective rule
that changes action across leaves. Between-leaf differences in $\Delta_L$
describe variation in magnitude, not the decision rule itself.

The standard public table therefore leads with `selected_action`,
`tc_score_contrast`, `score_interval`, `sample_percent`, and `direction`.
Lower-level diagnostics remain available, but selected-action advantage and
value-contribution columns are hidden unless explicitly requested.

The score interval is a source-specific action-score summary. For display-tree
tables, it is an approximate interval for row-level scores after tree selection.
For held-out CV tables, it summarises pooled held-out action-score summaries
from fold-level trees that may have different leaf structures. In neither case
should the interval be described as a formal post-selection subgroup test.

## Simulate two outcomes

The simulation creates two outcomes with related, but not identical,
heterogeneity. Age and socioeconomic position recur as candidate organising
variables, while the second outcome adds a distinct baseline variable.

```{r}
library(margot)
library(dplyr)

set.seed(20260620)
n <- 900

sim <- tibble(
  age_z = rnorm(n),
  status_z = rnorm(n),
  income_z = rnorm(n),
  baseline_y1 = rnorm(n),
  baseline_y2 = rnorm(n)
) |>
  mutate(
    propensity = plogis(-0.15 + 0.35 * age_z - 0.25 * status_z),
    exposure = rbinom(n(), 1, propensity),
    tau_y1 = 0.06 + 0.08 * (age_z > 0) + 0.04 * (status_z > 0),
    tau_y2 = 0.03 + 0.06 * (age_z > 0) - 0.05 * (income_z < -0.5),
    y1 = 0.25 * baseline_y1 + 0.15 * status_z + exposure * tau_y1 + rnorm(n(), sd = 0.8),
    y2 = 0.30 * baseline_y2 - 0.10 * income_z + exposure * tau_y2 + rnorm(n(), sd = 0.8)
  )

covariates <- sim |>
  select(age_z, status_z, income_z, baseline_y1, baseline_y2)
```

## Estimate outcome-wide ATEs

The ATE layer uses the fitted forests and `grf::average_treatment_effect()`.
Do not add an external policy-tree cross-validation step to the ATE estimate.

```{r}
fit <- margot_causal_forest(
  data = sim,
  outcome_vars = c("y1", "y2"),
  covariates = covariates,
  W = sim$exposure,
  weights = NULL,
  use_train_test_split = FALSE,
  compute_rate = FALSE,
  compute_conditional_means = FALSE,
  save_models = TRUE,
  save_data = TRUE,
  verbose = FALSE
)

ate_table <- margot_recompute_ate(fit)
ate_table
```

## Add bridge diagnostics

`grf::test_calibration()` evaluates whether forest predictions are calibrated on
held-out forest predictions. The differential-prediction coefficient is also an
omnibus diagnostic for heterogeneity. `grf::variable_importance()` is a
descriptive split-use summary. It should not be interpreted as a confirmed
moderator test.

```{r}
calibration <- lapply(fit$full_models, grf::test_calibration)

importance <- lapply(fit$full_models, function(forest) {
  tibble(
    variable = colnames(covariates),
    importance = as.numeric(grf::variable_importance(forest))
  ) |>
    arrange(desc(importance))
})

calibration
importance
```

## Evaluate policy trees on held-out folds

The policy-tree layer learns trees on training folds and evaluates the learned
tree on held-out folds. The output includes policy value, split frequencies,
threshold summaries, and leaf-level signed treatment-control contrasts.
Tree-level policy value is summarised against all-control, all-treatment, and
the best constant action, so an all-treatment or all-control tree is not
mistaken for a selective rule.

```{r}
policy_cv <- margot_policy_tree_cv(
  fit,
  verbose = FALSE
)

policy_cv$depth_selection
policy_cv$value_summary
margot_table_policy_value(policy_cv)
policy_cv$leaf_summary
```

The first call uses the workflow defaults. To make the defaults explicit in a
registration, record them as:

```{r}
# record the standard policy-tree validation settings in the study protocol.
policy_tree_settings <- list(
  depths = c(1L, 2L),
  num_folds = 5L,
  n_repeats = 20L,
  min_gain_for_depth_switch = 0.01,
  max_stability_loss_for_depth_switch = 0.05,
  tree_method = "fastpolicytree"
)
```

Users may restrict candidate policy-tree variables when the scientific question
justifies it. For confirmatory analyses, variable restrictions should be
pre-specified or chosen inside the training folds.

```{r}
policy_cv_subset <- margot_policy_tree_cv(
  fit,
  custom_covariates = c("age_z", "status_z", "income_z"),
  covariate_mode = "custom",
  verbose = FALSE
)
```

## Report a selected policy tree

The held-out CV object selects the reporting depth. The display tree then shows
the fitted full-sample tree at that depth. Display-tree leaf labels describe the
displayed tree; the held-out CV object remains the source for depth, value, and
split-frequency claims.

```{r}
selected_depth <- policy_cv$depth_map[["model_y1"]]

# use the policy-named wrapper for the branching decision tree.
margot_plot_policy_decision_tree(
  fit,
  model_name = "model_y1",
  max_depth = selected_depth,
  show_leaf_metrics = TRUE
)
```

Use `margot_plot_policy_projection()` when the intended artefact is the
evaluation-point projection, and `margot_plot_policy_tree_panels()` when the
report should show the branching tree in panel A and the projection in panel B.
The older `margot_plot_decision_tree()` and `margot_plot_policy_tree()` remain
available, but the policy-named wrappers make the intended plot type explicit.

The modular reporting helpers expose the same convention at different levels:

```{r}
# compute low-level leaf diagnostics for the selected display tree.
leaf_diagnostics <- margot_policy_leaf_summary(
  fit,
  model_name = "model_y1",
  depth = selected_depth
)

# create the public display-tree table with signed T-C first.
display_leaf_table <- margot_table_policy_tree(
  fit,
  model_name = "model_y1",
  depth = selected_depth
)

# create the held-out action-score summary table at the selected depth.
heldout_leaf_table <- margot_table_policy_tree(
  policy_cv,
  model_name = "model_y1",
  source = "heldout_cv"
)

# compare the learned policy with universal action baselines.
value_table <- margot_table_policy_value(
  policy_cv,
  model_name = "model_y1",
  depth = selected_depth
)

# generate cautious stock text for a manuscript or report.
policy_text <- margot_text_policy_tree(source = "heldout_cv")

display_leaf_table
heldout_leaf_table
value_table
policy_text
```

For exploratory diagnostics, callers can expose the lower-level selected-action
advantage and value-contribution columns. These columns are useful for auditing
the score arithmetic but should not lead public manuscript tables.

```{r}
# request diagnostic columns only when auditing the policy-tree arithmetic.
margot_table_policy_tree(
  fit,
  model_name = "model_y1",
  depth = selected_depth,
  include_selected_action_difference = TRUE,
  include_value_contribution = TRUE
)
```

The integrated helper assembles the standard plot, display-tree table,
held-out leaf table, held-out policy-value table, and interpretation text. Each
component remains an ordinary R object, so manuscript workflows can replace or
omit pieces without losing the shared reporting convention.

```{r}
# assemble the standard policy-tree report components for one outcome.
policy_report <- margot_report_policy_tree(
  fit,
  model_name = "model_y1",
  policy_cv = policy_cv
)

names(policy_report)
policy_report$metadata
policy_report$table
policy_report$heldout_table
policy_report$policy_value
policy_report$text
policy_report$plots$combined_plot
```

The table-level metadata flags whether the selected actions vary across leaves.
When `uniform_selected_action` is `TRUE`, avoid describing the tree as a
selective rule even if the tree has multiple leaves.

```{r}
# check whether the selected action changes across leaves.
display_leaf_table |>
  select(
    model,
    depth,
    node_id,
    selected_action,
    tc_score_contrast,
    n_selected_actions,
    uniform_selected_action
  )
```

The split summaries remain separate from leaf contrasts:

```{r}
# inspect root-split stability at the selected reporting depth.
policy_cv$split_summary |>
  filter(model == "model_y1", depth == selected_depth, node_id == 1)
```

## Summarise outcome-wide recurrence

Outcome-wide recurrence asks whether the same baseline variables recur across
outcomes. This layer is descriptive unless a study defines a formal family-level
target.

```{r}
recurrence <- margot_policy_recurrence_summary(policy_cv)
recurrence
```

A cautious report might say:

> Age recurred as a root or near-root policy-tree variable across both outcomes,
> while held-out value against the best constant action was small. We treat age
> as a recurring exploratory organiser, not a confirmed moderator.

## Optional extensions

RATE/AUTOC can be added when investigators need an explicit heterogeneity test.
When used, RATE/AUTOC should be cross-validated. Qini/uplift curves remain
optional and exploratory until the analysis has a clearly defined cost-benefit
interpretation.
