---
title: "Standard GRF and policy-tree workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard GRF and policy-tree workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette describes the standard `margot` workflow for generalised random
forest (GRF) analyses with policy-tree reporting. The workflow is designed for
outcome-wide studies: the same exposure, covariates, and analysis population are
used to estimate effects for multiple outcomes.

The workflow separates four tasks:

1. Estimate average treatment effects (ATEs) for each outcome.
2. Diagnose whether the forest predictions show calibrated heterogeneity.
3. Evaluate policy-tree learning with held-out folds.
4. Summarise cross-outcome recurrence descriptively.

The ATE layer estimates the primary causal estimand. The policy-tree layer asks
whether a shallow rule can summarise useful variation in the forest's doubly
robust action scores. Because a policy tree is an optimiser, policy-tree
learning should be evaluated on held-out observations.

## Why this simulation is simple

Athey, Imbens, Metzger, and Munro describe Wasserstein generative adversarial
networks (WGANs) for designing Monte Carlo simulations that mimic real economic
datasets. Their approach is valuable for estimator benchmarking, but it is not a
good default for a package tutorial because it adds Python, GAN training, large
generated-data artefacts, and additional tuning decisions. The standard `margot`
tutorial instead uses a transparent R simulation with known heterogeneity and
two outcomes. A WGAN benchmark could be added later as a separate stress-test
article.

## Simulate two outcomes

The simulation creates two outcomes with related, but not identical,
heterogeneity. Age and socioeconomic position recur as candidate organising
variables, while the second outcome adds a distinct baseline variable.

```{r}
library(margot)
library(dplyr)

set.seed(20260620)
n <- 900

sim <- tibble(
  age_z = rnorm(n),
  status_z = rnorm(n),
  income_z = rnorm(n),
  baseline_y1 = rnorm(n),
  baseline_y2 = rnorm(n)
) |>
  mutate(
    propensity = plogis(-0.15 + 0.35 * age_z - 0.25 * status_z),
    exposure = rbinom(n(), 1, propensity),
    tau_y1 = 0.06 + 0.08 * (age_z > 0) + 0.04 * (status_z > 0),
    tau_y2 = 0.03 + 0.06 * (age_z > 0) - 0.05 * (income_z < -0.5),
    y1 = 0.25 * baseline_y1 + 0.15 * status_z + exposure * tau_y1 + rnorm(n(), sd = 0.8),
    y2 = 0.30 * baseline_y2 - 0.10 * income_z + exposure * tau_y2 + rnorm(n(), sd = 0.8)
  )

covariates <- sim |>
  select(age_z, status_z, income_z, baseline_y1, baseline_y2)
```

## Estimate outcome-wide ATEs

The ATE layer uses the fitted forests and `grf::average_treatment_effect()`.
Do not add an external policy-tree cross-validation step to the ATE estimate.

```{r}
fit <- margot_causal_forest(
  data = sim,
  outcome_vars = c("y1", "y2"),
  covariates = covariates,
  W = sim$exposure,
  grf_defaults = list(num.trees = 500, min.node.size = 20, seed = 42),
  use_train_test_split = FALSE,
  compute_rate = FALSE,
  compute_conditional_means = FALSE,
  save_models = TRUE,
  save_data = TRUE,
  verbose = FALSE
)

ate_table <- margot_recompute_ate(fit)
ate_table
```

## Add bridge diagnostics

`grf::test_calibration()` evaluates whether forest predictions are calibrated on
held-out forest predictions. The differential-prediction coefficient is also an
omnibus diagnostic for heterogeneity. `grf::variable_importance()` is a
descriptive split-use summary. It should not be interpreted as a confirmed
moderator test.

```{r}
calibration <- lapply(fit$full_models, grf::test_calibration)

importance <- lapply(fit$full_models, function(forest) {
  tibble(
    variable = colnames(covariates),
    importance = as.numeric(grf::variable_importance(forest))
  ) |>
    arrange(desc(importance))
})

calibration
importance
```

## Evaluate policy trees on held-out folds

The policy-tree layer learns trees on training folds and evaluates the learned
tree on held-out folds. The output includes policy value, split frequencies,
threshold summaries, and leaf-level estimated gains.

```{r}
policy_cv <- margot_policy_tree_cv(
  fit,
  depths = c(1, 2),
  num_folds = 5,
  n_repeats = 10,
  min_gain_for_depth_switch = 0.01,
  max_stability_loss_for_depth_switch = 0.05,
  tree_method = "fastpolicytree",
  seed = 2026,
  verbose = FALSE
)

policy_cv$depth_selection
policy_cv$value_summary
policy_cv$leaf_summary
```

Users may restrict candidate policy-tree variables when the scientific question
justifies it. For confirmatory analyses, variable restrictions should be
pre-specified or chosen inside the training folds.

```{r}
policy_cv_subset <- margot_policy_tree_cv(
  fit,
  custom_covariates = c("age_z", "status_z", "income_z"),
  covariate_mode = "custom",
  depths = c(1, 2),
  num_folds = 5,
  n_repeats = 10,
  seed = 2026,
  verbose = FALSE
)
```

## Plot selected display trees

The plot below shows a stored tree and annotates leaves with action-conditional
estimated gains and sample shares. These labels describe the displayed tree. The
held-out CV object remains the source for depth, value, and split-frequency
claims.

```{r}
selected_depth <- policy_cv$depth_map[["model_y1"]]

margot_plot_decision_tree(
  fit,
  model_name = "model_y1",
  max_depth = selected_depth,
  show_leaf_metrics = TRUE
)
```

For a report, pair the plot with held-out summaries:

```{r}
policy_cv$split_summary |>
  filter(model == "model_y1", depth == selected_depth, node_id == 1)

policy_cv$leaf_summary |>
  filter(model == "model_y1", depth == selected_depth)
```

## Summarise outcome-wide recurrence

Outcome-wide recurrence asks whether the same baseline variables recur across
outcomes. This layer is descriptive unless a study defines a formal family-level
target.

```{r}
recurrence <- margot_policy_recurrence_summary(policy_cv)
recurrence
```

A cautious report might say:

> Age recurred as a root or near-root policy-tree variable across both outcomes,
> but held-out gains were small. We treat age as a recurring exploratory
> organiser, not a confirmed moderator.

## Optional extensions

RATE/AUTOC can be added when investigators need an explicit heterogeneity test.
When used, RATE/AUTOC should be cross-validated. Qini/uplift curves remain
optional and exploratory until the analysis has a clearly defined cost-benefit
interpretation.
